### 2017-07-15 Blog/07-WhatIsInformationEntropy.tex
Final edit to entropy blog post

TL;DR Information entropy tells you the average number of bits (i.e. answers to yes/no questions) you need to identify an element from a distribution.

@@

The concept of entropy in physics is unfortunately associated with a certain amount of vagueness. While mathematically is well defined, I have never heard a crisp conceptual understanding (e.g. it measures "disorder", it has to do with how much the system is homogeneous, ...). Entropy in information theory, or Shannon's entropy, instead has a crisp intuitive meaning and therefore I cringe when I hear people describing it as "lack of information" or any other vague description. So, let's see if we can do better.

The scenario is this: you have a distribution over a set of possibilities. You pick an element according to that distribution. On average, how many bits of information you have to use to tell me which element have you picked? How many one-zeros? How many yes-no answers? If that doesn't make sense yet, hopefully it will after we go through a few examples.

@@

Suppose you have a bag with forty balls. Twenty are green and twenty are red. Suppose you pick a ball and I want to know the color. Suppose I can only ask you yes or no questions. How many questions do I have to ask you? One suffices: "is it red?". If the answer is yes, I know it's red, if the answer is no, I know it's green. We can rephrase by asking: how many one-zero choices? That is: how many bits of information you need to give me? The answer is one bit: 0 for green and 1 for red. That is: all the information can be encoded into a single zero-one choice.

@@

Suppose there are ten balls for each of these color: green, red, yellow and blue. How many bits of information? We can set 00 for green, 01 for red, 10 for yellow and 11 for blue. Four cases can be encoded into two zero-one choices: it's two bits of information.

Shannon's entropy simply generalizes this idea: given a $n$ cases each with a probability $p_i$, the average number of bits (i.e. yes/no answers) required to identify an element is:

@@

The base of the logarithm determines the units. If it's base $2$, then we are using bits. But sometimes is useful to use base $e$ (i.e. the natural log) since it has nice properties. In that case we are using nats, and we have $1$ bit = $\ln 2$ nats.

@@

\begin{enumerate}
	\item $H(p_1, ..., p_n)$ is continuous.
	\item $H(p_1, ..., p_n)$ increases if the number of choice increases. That is, suppose you have two uniform distributions over $n$ and $m$ cases respectively. If $m>n$ then $H(\frac{1}{m}, ..., \frac{1}{m})>H(\frac{1}{n}, ..., \frac{1}{n})$.
	\item $H(p_1, ..., p_n)$ does not change if we group/break down choices. Consider the following trees:
	\begin{figure}[h]
	\includegraphics[scale=0.50]{ShannonTree}
	\centering
	\end{figure}

@@

While we are not going to derive it here, the derivation is not particularly insightful, we can still understand why it is what it is. If a particular case is a choice among $n$ equal cases, it's probability is $p_i=\frac{1}{n}$ and the number of yes/no questions we have to make to get to it is $\log_2 n = \log_2 \frac{1}{p_i} = - \log_2 p_i$. So, when that particular case is selected, we need $- \log_2 p_i$ bits. Now we take the expectation, the average over all cases, and we have $H(p_1, ..., p_n) = - \sum_{i=1}^{n} p_i \log p_i$. We now have the average value of yes/no questions (i.e. bits) needed to identify an element in the distribution.

@@

There are a couple of aspects we need to clarify as it allows to understand why such concept is useful in physics.

First of all, information entropy has nothing to do with knowledge (e.g. what we know about the system or what can be known). This misunderstanding arises because the concept can be applied to probability distribution that represent our knowledge of the system, and in some cases it may relate to knowledge. For example, if there are four boxes and I do not know in which one the ball is hiding, I have a uniform distribution on four cases which has 2 bits of information entropy. Conversely, if I know where it is, the entropy is 0 bits. So, in this case more entropy correspond to less knowledge. But this idea does not generalize.

@@

Moreover, not all distributions represent knowledge and/or are statistical. For example, we could have a mass distribution in space. In this case, information entropy does not equate to lack of knowledge. Our knowledge of the mass distribution is perfect: we know exactly how it is spread over a region. The information entropy of a mass distribution in space is as much a physical property of the system as it is the mass distribution itself and it tells us how many bits are necessary to identify the position of a part of the system. Unlike the previous case, this is an objective number. It is this second use that is the most relevant to physics. Nature does not much care about our knowledge: it cares about objective facts.

@@

At the end of the day, information entropy is a number: to understand it you need to understand what that number quantifies and in what units. If do not have a clear idea what that number is measuring, you do not understand what it is. That number measures how many bits of informations are required on average to identify an element in a distribution. Not "knowledge" or "lack of information" or "disorder" or any other vague concept.

>>>>

TL;DR Information entropy tells you the average number of bits (i.e.~answers to yes/no questions) you need to identify an element from a distribution.

@@

The concept of entropy in physics is unfortunately associated with a certain amount of vagueness. While mathematically it is well defined, I have never heard a crisp conceptual understanding (e.g. it measures ``disorder", it has to do with how much the system is homogeneous, something to do with the number of microstates, ...). Entropy in information theory, or Shannon's entropy, instead has a crisp intuitive meaning and therefore I cringe when I hear people describing it as ``lack of information" or any other vague description. So, let's see if we can do better.

The scenario is this: you have a distribution over a set of possibilities. You pick an element according to that distribution. On average, how many bits of information do you have to use to tell me which element you have picked? How many one-zeros? How many yes-no answers? If that doesn't make sense yet, hopefully it will after we go through a few examples.

@@

Suppose you have a bag with forty balls. Twenty are green and twenty are red. Suppose you pick a ball and I want to know the color. Suppose I can only ask you yes or no questions. How many questions do I have to ask you? One suffices: ``Is it red?" If the answer is yes, I know it's red, if the answer is no, I know it's green. We can rephrase by asking: how many one-zero choices? That is: how many bits of information you need to give me? The answer is one bit: 0 for green and 1 for red. That is: all the information can be encoded into a single zero-one choice.

@@

Suppose there are ten balls for each of these colors: green, red, yellow and blue. How many bits of information? We can set 00 for green, 01 for red, 10 for yellow and 11 for blue. Four cases can be encoded into two zero-one choices: it's two bits of information.

Shannon's entropy simply generalizes this idea: given $n$ cases each with a probability $p_i$, the average number of bits (i.e. yes/no answers) required to identify an element is:

@@

The base of the logarithm determines the units. If it's base $2$, then we are using bits. But sometimes it is useful to use base $e$ (i.e.~the natural log) since it has nice properties. In that case we are using nats (as in natural log), and we have $1$ bit = $\ln 2$ nats.

@@

\begin{enumerate}
	\item $H(p_1, ..., p_n)$ is continuous.
	\item $H(p_1, ..., p_n)$ increases if the number of choices increases. That is, suppose you have two uniform distributions over $n$ and $m$ cases respectively. If $m>n$ then $H(\frac{1}{m}, ..., \frac{1}{m})>H(\frac{1}{n}, ..., \frac{1}{n})$.
	\item $H(p_1, ..., p_n)$ does not change if we group/break down choices. Consider the following trees:
	\begin{figure}[h]
	\includegraphics[scale=0.50]{ShannonTree}
	\centering
	\end{figure}

@@

Since the derivation is not particularly insightful, we are not going to derive it here. But we can still understand why it is what it is. If a particular case is a choice among $n$ equal cases, its probability is $p_i=\frac{1}{n}$ and the number of yes/no questions we have to make to get to it is $\log_2 n = \log_2 \frac{1}{p_i} = - \log_2 p_i$. So, when that particular case is selected, we need $- \log_2 p_i$ bits. Now we take the average over all cases and we have $H(p_1, ..., p_n) = \sum_{i=1}^{n} p_i \, (- \log_2 p_i)$. We now have the average number of yes/no questions (i.e.~bits) needed to identify an element in the distribution.

@@

There are a couple of aspects we need to clarify to understand why this concept is useful in physics.

First of all, information entropy has nothing to do with knowledge (e.g.~what we know about the system or what can be known). This misunderstanding arises because the concept can be applied to probability distributions that represent our knowledge of the system, and in some cases it may seem to relate to knowledge. For example, if there are four boxes and I do not know in which one the ball is hiding, I have a uniform distribution on four cases which has 2 bits of information entropy. Conversely, if I know where it is, the entropy is 0 bits. So, in this case more entropy corresponds to less knowledge. But this is purely coincidental.

@@

Moreover, not all distributions represent knowledge and/or are statistical. For example, we could have a mass distribution in space. In this case, information entropy clearly cannot correspond to lack of knowledge. Our knowledge of the mass distribution is perfect: we know exactly how it is spread over a region. The information entropy of a mass distribution in space is as much a physical property of the system as is the mass distribution itself and it tells us how many bits are necessary to identify the position of a part of the system. Unlike the previous case, this is an objective number. It is this second use that is the most relevant to physics. Nature does not much care about our knowledge: it cares about objective facts.

@@

At the end of the day, information entropy is a number: to understand it you need to understand what that number quantifies and in what units. If you do not have a clear idea what that number is measuring, you do not understand what it is. That number measures how many bits of informations are required on average to identify an element in a distribution. Not ``knowledge" or ``lack of information" or ``disorder" or any other vague concept.
